---
title: "Get Started with landscapemetrics"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    html_preview: false
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load_libraries_hidden, eval=TRUE, echo=FALSE, message=FALSE, results='hide'}
library(landscapemetrics)
library(raster)
library(dplyr)
```

# Using **landscapemetrics**

All functions in **landscapemetrics** start with `lsm_` (for landscapemetrics). The second part of the name specifies the level (patch - `p`, class - `c` or landscape - `l`). The last part of the function name is the abbreviation of the corresponding metric (e.g. `enn`for the euclidean nearest-neighbor distance):

```
# general structure
lsm_"level"_"metric"

# Patch level
## lsm_p_"metric"
lsm_p_enn()

# Class level
## lsm_c_"metric"
lsm_c_enn()

# Landscape level
## lsm_p_"metric"
lsm_l_enn()
```

All functions return an identical structured tibble:

<p style="text-align:center;">

| layer  | level | class | id | metric | value
| ------------- | -------------  | ------------- | ------------- | ------------- | ------------- | 
| 1 | patch  | 1 | 1 | landscape metric | x |
| 1 | class  | 1 | NA | landscape metric | x |
| 1 | landscape  | NA | NA | landscape metric | x |

</p>

```{r, message=FALSE}
# Import raster
landscape_raster <- landscape
# for local file: raster("pathtoyourraster/raster.asc")
# ... or any other raster file type, geotiff, ...

# Calculate e.g. perimeter of all patches
lsm_p_perim(landscape_raster)
```

The metrics are encoded using abbreviations in the result tibble. In case you want to match the full name of each metric, we provide a tibble (`lsm_abbreciations_names`) you can join with the result tibble using the `metric` column. The this also adds the type of the metrics.

```{r, message=FALSE}
# Calculate metric
result <- lsm_c_dcad(landscape)

# Left join with abbreviation tibble
result_full_name <- left_join(x = result, 
                                     y = lsm_abbreviations_names, 
                                     by = "metric")
# Show results
result_full_name
```

## Important information about using **landscapemetrics**

The resolution of a raster cell has to be in **meters**, as the package converts units internally and returns results in either meters, square meters or hectares. For more information see the help file of each function.

## Using **landscapemetrics** in a tidy workflow

Every function in *landscapemetrics* accept data as its first argument, which makes piping a natural workflow. A possible use case is that you would load your spatial data, calculate some landscape metrics and then use the resulting tibble in further analyses.

```{r, message=FALSE}
# All patch IDs of class 2 with an ENN > 2.5
subsample_patches <- landscape %>% 
    lsm_p_enn() %>%
    dplyr::filter(class == 2 & value > 2.5) %>%
    dplyr::pull(id)

# Show results
subsample_patches
``` 

### Use multiple metric functions

As the result of every function always returns a tibble, combining the metrics that were selected for your research question is straightforward:

```{r, message=FALSE}
# bind results from different metric functions
patch_metrics <- bind_rows(
  lsm_p_cai(landscape),
  lsm_p_circle(landscape),
  lsm_p_enn(landscape)
  )
# look at the results
patch_metrics 
```

Additionally, we provide a wrapper where the desired metrics can be specified as a vector of strings. Because all metrics regardless of the level return an identical tibble, different levels can be mixed. It is also possible to calculate all available metrics at a certain level using e.g. `what = "patch"`

```{r, message=FALSE}
calculate_metrics(landscape, what = c("lsm_c_pland", "lsm_l_ta", "lsm_l_te"))
```

All metrics are abbreviated in the result `tibble`. Therefore, we provide a `tibble` containing the full metric names, as well as the class of each metric (`lsm_abbreviations_names`). Using e.g. the `left_join()` function of the `dplyr` package one could join a result `tibble` and the abbrevations `tibble`.

```{r, message=FALSE}

# bind results from different metric functions
patch_metrics <- bind_rows(
  lsm_p_cai(landscape),
  lsm_p_circle(landscape),
  lsm_p_enn(landscape)
  )
# look at the results
patch_metrics_full_names <- dplyr::left_join(x = patch_metrics,
                                             y = lsm_abbreviations_names, 
                                             by = "metric")
patch_metrics_full_names
```

In `calculate_metrics` this exists as option (`full_name = TRUE`).

```{r}
calculate_metrics(landscape, what = c("lsm_c_pland", "lsm_l_ta", "lsm_l_te"), 
              full_name = TRUE)

```
