---
title: "Differences FRAGSTATS|landscapemetrics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differences FRAGSTATS|landscapemetrics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
```{r load_libraries_hidden, message = FALSE, warning = FALSE, include = FALSE}
library(DT)
library(landscapemetrics)
library(landscapetools)
library(patchwork)
library(raster)
library(tidyverse)
```

**landscapemetrics** re-implements landcsape metrics as they are described in the **FRAGSTATS** software (McGarigal et al. 2012). Therefore, we compared our results with the results of **FRAGSTATS**. In the process, we recognized a few differences between the results. 

Some metrics in *FRAGSTATS* are interdependent across scales. Thus, if there is a deviation at the patch level, it propagates through the class- and landscape-level. We list the metrics with deviations at the lowest level.

Unfortunatly, we do not have access to the source code of **FRAGSTATS**. Therefore, we are not able to finally explain the present differences between the results, nevertheless, we try to guess the most likely reasons.

# General differences

Firstly, the patch ID is ordered in a different way, most likely due to technical reasons (how connected patches are specified). Therefore, one has to pay attention comparing the results on patch level for **FRAGSTATS** and **landscapemetrics**.

All double precision floating point numbers are rounded after the 4th decimal place in **FRAGSTATS**. Contrastingly, we do not round the numbers. Naturally, this can lead to small deviations between the results.

There are quite a few metrics on class- and landscape-level that summarise patch level metrics (e.g. the mean, standard deviation (sd) or coefficient of variation (cv) of all values belonging to class i). While the results are identical for the single patches and the mean of all patches, there are some slight differences between **lanscapemetrics** and **FRAGSTATS** for the standard deviation and the coefficent of variation.

In the following, we are comparing the cv for the patch area. We are including the cv calculate from all patch areas and the acutal output of **FRAGSTATS** as well as the output of **landscapemetrics**. Interestingly, the cv calculated from all patches of **FRAGSTATS** is identical to the cv of **landscapemetrics**, but the actual result of **FRAGSTATS** is slightly different.

```{r}
# function to calculate coefficient of variation
cv <- function(x) {
    (sd(x, na.rm = TRUE) /
         mean(x, na.rm = TRUE)) * 100
}

# CV calculated from patch values of FRAGSTATS 
fragstats_calculated <- fragstats_patch_landscape %>%
    dplyr::group_by(TYPE) %>%
    dplyr::summarise(cv = cv(AREA)) %>%
    purrr::set_names("class", "fragstats_calculated") %>%
    dplyr::mutate(class = as.integer(dplyr::case_when(class == "cls_1" ~ 1,
                                                      class == "cls_2" ~ 2,
                                                      class == "cls_3" ~ 3)))

# Output of FRAGSTATS
fragstats_output <- fragstats_class_landscape %>% 
    dplyr::select(TYPE, AREA_CV) %>%
    purrr::set_names("class", "fragstats_output") %>%
    dplyr::mutate(class = as.integer(dplyr::case_when(class == "cls_1" ~ 1,
                                                      class == "cls_2" ~ 2,
                                                      class == "cls_3" ~ 3)))

# Output of landscapemetrics
landscapemetrics_output <- lsm_c_area_cv(landscape) %>%
    dplyr::select(class, value) %>%
    purrr::set_names("class", "landscapemetrics")
```

```{r echo = FALSE}
fragstats <- dplyr::full_join(x = fragstats_calculated, 
                              y = fragstats_output, 
                              by = "class") 

cv_full <- dplyr::full_join(x = fragstats, 
                            y = landscapemetrics_output, 
                            by = "class")

print.data.frame(cv_full)
```

As for the cv, the results for the sd are similiar. The result calculated from all patch areas of **FRAGSTATS** is identical to the result of **landscapemetrics**, but not the actual result of **FRAGSTATS**.

```{r}
# SD calculated from patch values of FRAGSTATS
fragstats_calculated <- fragstats_patch_landscape %>%
    group_by(TYPE) %>%
    summarise(sd = sd(AREA)) %>%
    purrr::set_names("class", "fragstats_calculated") %>%
    dplyr::mutate(class = as.integer(dplyr::case_when(class == "cls_1" ~ 1,
                                                      class == "cls_2" ~ 2,
                                                      class == "cls_3" ~ 3)))

    
# Output of FRAGSTATS
fragstats_output <- fragstats_class_landscape %>%
    dplyr::select(TYPE, AREA_SD) %>%
    purrr::set_names("class", "fragstats_calculated") %>%
    dplyr::mutate(class = as.integer(dplyr::case_when(class == "cls_1" ~ 1,
                                                      class == "cls_2" ~ 2,
                                                      class == "cls_3" ~ 3)))


# Output of landscapemetrics
landscapemetrics_output <- lsm_c_area_sd(landscape) %>%
    dplyr::select(class, value) %>%
    purrr::set_names("class", "landscapemetrics")
    
```

```{r echo = FALSE}
fragstats <- dplyr::full_join(x = fragstats_calculated, 
                              y = fragstats_output, 
                              by = "class") 

cv_full <- dplyr::full_join(x = fragstats, 
                            y = landscapemetrics_output, 
                            by = "class")

print.data.frame(cv_full)
```

# Specific differences

### CIRCLE metric

**FRAGSTATS** defines CIRCLE as following:

> [...] CIRCLE equals 1 minus patch area (m^2^) divided by the area (m^2^) of the smallest circumscribing circle. [...]

According to the definition, for a squared patch the metric should equal CIRCLE = 0.3633802, regardless of the actual patch size (e.g. 1 cell vs. 4-by-4 cells). However, in **FRAGSTATS** patches containing only one cell equal CIRCLE = 0. In the following, patches with the area of area = 0.0001 contain only 1 cell and are following a perfect square.

```{r}
# Calculate patch area
fragstats_area <- fragstats_patch_landscape %>%
    select(PID, AREA) %>%
    purrr::set_names("id", "fragstats_area")

landscapemetrics_area <- lsm_p_area(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "landscapemetrics_area")

# Calculate CIRCLE
fragstats_circle <- fragstats_patch_landscape %>%
    select(PID, CIRCLE) %>%
    purrr::set_names("id", "fragstats_circle")

landscapemetrics_circle <- lsm_p_circle(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "landscapemetrics_circle")
```

```{r echo = FALSE}
fragstats <- dplyr::full_join(x = fragstats_area, 
                              y = fragstats_circle, 
                              by = "id") %>%
    dplyr::arrange(fragstats_area)

landscapemetrics <- dplyr::full_join(x = landscapemetrics_area, 
                                     y = landscapemetrics_circle, 
                                     by = "id") %>%
    dplyr::arrange(landscapemetrics_area)

print.data.frame(fragstats)
print.data.frame(landscapemetrics)
```

#### Error propagation (for metrics based on **CIRCLE** metric)

##### Class level

- CIRCLE_CV (`lsm_c_circle_cv`)
- CIRCLE_MN (`lsm_c_circle_mn`)
- CIRCLE_SD (`lsm_c_circle_sd`)

##### Landscape level

- CIRCLE_CV (`lsm_l_circle_cv`)
- CIRCLE_MN (`lsm_l_circle_mn`)
- CIRCLE_SD (`lsm_l_circle_sd`)

##### Landscape level

### CORE metric

In **FRAGSTATS** the core metric is defined as:

> [...] CORE equals the area (m^2^) within the patch that is further than the specified depth-of-edge distance from the patch perimeter, [...]

In **landscapemetrics** the depth-of-edge is fixed to one cell. In other words, all cells with four adjacent neighbours of the same class are defined as core cells. The implementation of the depth-of-edge distance seems to be slightly different in **FRAGSTATS**, leading to slightly different results for core metrics. 

This is a visual aid to help understand how we implemented the core metric. The left plot shows all patches of class 1, the right plot the core area (violet cells) in each patch.

```{r}
# All patches of class 1
class_1 <- cclabel(landscape)[[1]]

# Core area of all patches, 1 = Edge, 0 = Core
core <- boundaries(class_1, directions = 4)
```

```{r echo = FALSE, message = FALSE, fig.align = 'center', fig.retina = 2}
# Plot all patches of class 1 (left) and core area (right)
util_plot(class_1) + util_plot(core)
```

#### Error propagation (for metrics based on **CORE** metric)

##### Patch level

- NCORE metric (`lsm_p_ncore`)
- CAI metric (`lsm_p_cai`)

##### Class level

- CAI_CV (`lsm_c_cai_cv`)
- CAI_MN (`lsm_c_cai_mn`)
- CAI_SD (`lsm_c_cai_sd`)
- CORE_CV (`lsm_c_core_cv`)
- CORE_MN (`lsm_c_core_mn`)
- CORE_SD (`lsm_c_core_sd`)
- CPLAND (`lsm_c_cpland`)
- DCAD (`lsm_c_dcad`)
- DCORE_CV (`lsm_c_dcore_cv`)
- DCORE_MN (`lsm_c_dcore_mn`)
- DCORE_SD (`lsm_c_dcore_sd`)
- NDCA (`lsm_c_ndca_sd`)
- TCA (`lsm_c_tca`)

##### Landscape level

- CAI_CV (`lsm_l_cai_cv`)
- CAI_MN (`lsm_l_cai_mn`)
- CAI_SD (`lsm_l_cai_sd`)
- CORE_CV (`lsm_l_core_cv`)
- CORE_MN (`lsm_l_core_mn`)
- CORE_SD (`lsm_l_core_sd`)
- DCAD (`lsm_l_dcad`)
- DCORE_CV (`lsm_l_dcore_cv`)
- DCORE_MN (`lsm_l_dcore_mn`)
- DCORE_SD (`lsm_l_dcore_sd`)
- NDCA (`lsm_l_ndca`)
- TCA (`lsm_l_ndca`)

### GYRATE metric

According to **FRAGSTATS** the radius of gyration for a patch consisting of only a single cell should equal GYRATE = 0.

> [...] GYRATE = 0 when the patch consists of a single cell [...]

However, for patches containing a single cell **FRAGSTATS** returns a value of GYRATE = 0.5. In the following table, patches with an area of area = 0.0001 consist of only one cell. 

```{r}
# Calculate patch area
fragstats_area <- fragstats_patch_landscape %>%
    select(PID, AREA) %>%
    purrr::set_names("id", "fragstats_area")

landscapemetrics_area <- lsm_p_area(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "landscapemetrics_area")

# Calculate GYRATE
fragstats_circle <- fragstats_patch_landscape %>%
    select(PID, GYRATE) %>%
    purrr::set_names("id", "fragstats_gyrate")

landscapemetrics_circle <- lsm_p_gyrate(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "landscapemetrics_gyrate")
```

```{r echo = FALSE}
fragstats <- dplyr::full_join(x = fragstats_area, 
                              y = fragstats_circle, 
                              by = "id") %>%
    dplyr::arrange(fragstats_area)

landscapemetrics <- dplyr::full_join(x = landscapemetrics_area, 
                                     y = landscapemetrics_circle, 
                                     by = "id") %>%
    dplyr::arrange(landscapemetrics_area)

print.data.frame(fragstats)
print.data.frame(landscapemetrics)
```

#### Error propagation (for metrics based on **GYRATE** metric)

##### Class level

- GYRATE_CV (`lsm_c_gyrate_cv`)
- GYRATE_MN (`lsm_c_gyrate_mn`)
- GYRATE_SD (`lsm_c_gyrate_sd`)

##### Landscape level

- GYRATE_CV (`lsm_l_gyrate_cv`)
- GYRATE_MN (`lsm_l_gyrate_mn`)
- GYRATE_SD (`lsm_l_gyrate_sd`)

### PARA metric

The documentation of **FRAGSTATS** defines the perimeter-area ratio the following: 

> [...] PARA equals the ratio of the patch perimeter (m) to area (m2). [...]

Contrastingly, the output of **FRAGSTATS** gives the result as the ratio of the patch perimeter in meters to area in hectares.

We implemented PARA as documented in the **FRAGSTATS** manual using square meters. Nevertheless, the differences between the softwares are only based on different units, as shown by converting the **FRAGSTATS** output to meters per square meters. 

```{r}
# Output of FRAGSTATS
fragstats <- fragstats_patch_landscape %>%
    dplyr::select(PID, AREA, PERIM, PARA) %>%
    purrr::set_names("id", "area", "perim", "para") %>%
    dplyr::mutate(para_calculated_ha = perim / area, 
                  para_calculated_m = perim / (area * 10000)) %>%
    dplyr::arrange(area)

# Output of landscapemetrics
area_landscapmetrics <- lsm_p_area(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "area")

perim_landscapmetrics <- lsm_p_perim(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "perim")

para_landscapemetrics <- lsm_p_para(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "para")

landscapemetrics <- dplyr::full_join(x = area_landscapmetrics, 
                                     y = perim_landscapmetrics, 
                                     by = "id") %>%
    dplyr::full_join(para_landscapemetrics, 
                     by = "id") %>%
    dplyr::mutate(para_calculated_ha = perim / area, 
                  para_calculated_m = perim / (area * 10000)) %>% 
    dplyr::arrange(area)
```

```{r echo = FALSE}
print.data.frame(fragstats)
print.data.frame(landscapemetrics)
```

#### Error propagation (for metrics based on **PARA** metric)

##### Class level
- PARA_MN (`lsm_c_para_mn`)
- PARA_SD (`lsm_c_para_sd`)

### References 

McGarigal, K., SA Cushman, and E Ene. 2012. FRAGSTATS v4: Spatial Pattern Analysis Program for Categorical and Continuous Maps. Computer software program produced by the authors at the University of Massachusetts, Amherst. Available at the following website: http://www.umass.edu/landeco/research/fragstats/fragstats.html
