---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r load_libraries_hidden, eval=TRUE, echo=FALSE, message=FALSE, results='hide'}
library(landscapemetrics)
library(tidyverse)
library(DT)
```

## Known issues

- FRAGSTATS only rounds after the 4th decimal place
   - That also effects the resolution, if your raster has a resolution with more 
     than 4 decimal places the results for metrics that rely on that are wrong (e.g. AREA, xxx) 

### Possible errors in FRAGSTATS for certain metrics

#### CIRCLE metric
Patches with 1 cell get a CIRCLE metric of 0, should be something around 0.36 instead.
Sometimes the abbreviation is CIRCLE, sometimes SQUARE.

#### PARA metric
There is a mixup between hectare and meter in documentation and code of FRAGSTATS.
The program uses hectare, whereas the documentation states in the formula square meter.

#### SHAPE metric
The documentation of FRAGSTATS formulates the SHAPE metric on patch level as:

\deqn{(0.25 * PERIM) / sqrt(AREA)}

... and FRAGSTATS give as result for a patch with 5 cells and a perimeter of 12 
the value of 1.200.

However, implementing the same formula in R results in:

```{r}
(0.25 * 12) / sqrt(5)
```

## Patch level

## Class level

```{r, echo=FALSE}
fs_class = data.frame(
    metric = c(
        "total area",
        "patch area (mean)",
        "patch area (mean)",
        "patch area (mean)",
        "patch area (cv)",
        "patch area (cv)",
        "patch area (cv)",
        "patch area (sd)",
        "patch area (sd)",
        "patch area (sd)",
        "percentage of landscape",
        "percentage of landscape",
        "percentage of landscape",
        "largest patch index",
        "largest patch index",
        "largest patch index",
        "number of patches",
        "number of patches",
        "number of patches",
        "euclidean nearest neighbor distance distribution (mean)",
        "euclidean nearest neighbor distance distribution (mean)",
        "euclidean nearest neighbor distance distribution (mean)"
    ),
    class = c(NA, 1, 3, 2, 1, 3, 2, 1, 3, 2, 1, 3, 2, 1, 3, 2,
              1, 3, 2, 1, 3, 2),
    value_fs = c(
        NA,
        0.002,
        0.012,
        0.0017,
        228.5906,
        162.6136,
        159.0109,
        0.0045,
        0.0195,
        0.0027,
        19.8889,
        53.2222,
        26.8889,
        16.4444,
        50.7778,
        10.8889,
        9,
        4,
        14,
        3.6829,
        2,
        3.1969
    )
)

lsm_class = lsm_calculate(landscape, what = "class")

class = left_join(fs_class, lsm_class, by = c("metric", "class")) %>% 
    select(starts_with("valu")) %>% 
    select(Value_FRAGSTATS = value_fs, Value_landscapemetrics = value)

datatable(class)
```

## Landscape level

```{r, echo=FALSE}
fs_land = data.frame(
    metric = c(
        "total area",
        "patch area (mean)",
        "patch area (cv)",
        "patch area (sd)",
        "largest patch index",
        "total edge",
        "number of patches",
        "patch richness",
        "patch richness density",
        "relative patch richness",
        "euclidean nearest neighbor distance distribution (mean)",
        "Shannon's evenness index",
        "Shannon's diversity index"
    ),
    value_fs = c(
        0.09,
        0.0033,
        268.3046,
        0.0089,
        50.7778,
        364,
        27,
        3,
        3333.3333,
        NA,
        .1816,
        0.9194,
        1.0101
    )
)

lsm_land = lsm_calculate(landscape, what = "landscape")

land = left_join(fs_land, lsm_land, by = "metric") %>%
    select(starts_with("valu")) %>% 
    select(Value_FRAGSTATS = value_fs, Value_landscapemetrics = value)

datatable(land)
```
