---
title: "Differences FRAGSTATS|landscapemetrics|SDMTools"
date: "`r Sys.Date()`"
output: 
  github_document:
    html_preview: false
---
```{r load_libraries_hidden, message = FALSE, warning = FALSE, include = FALSE}
library(landscapemetrics)
library(landscapetools)
library(patchwork)
library(raster)
library(tidyverse)
```

# Comparison with FRAGSTATS

`landscapemetrics` re-implements landscape metrics as they are mostly described in the FRAGSTATS software (McGarigal *et al.* 2012). Therefore, we compared our results with the results of FRAGSTATS. In the process, we recognized a few differences between the results. 

Some metrics in FRAGSTATS are interdependent across scales.  Thus, if there is a deviation at the patch level, it propagates through the class- and landscape-level. We list the metrics with deviations at the lowest level.

Unfortunatly, we do not have access to the source code of FRAGSTATS.  Therefore, we are not able to finally explain the present differences between the results, nevertheless, we try to guess the most likely reasons.

## General differences

Firstly, the patch ID is ordered in a different way, due to technical reasons (how connected patches are specified). Therefore, one has to pay attention comparing the results on patch level for FRAGSTATS and `landscapemetrics`.

All double precision floating point numbers are rounded after the 4th decimal place in FRAGSTATS. Contrastingly, we do not round the numbers. Naturally, this can lead to small deviations between the results.

There are quite a few metrics on class- and landscape-level that summarise patch level metrics (e.g. the mean, standard deviation (*sd*) or coefficient of variation (*cv*) of all values belonging to class *i*). While the results are identical for single patches and the mean of all patches, there are some slight differences between `lanscapemetrics` and FRAGSTATS for the standard deviation and the coefficent of variation. `landscapemetrics` uses base `R` functions for that, so we should assume that the calculation of such indices is correct.

In the following, we are comparing the *cv* for the patch area. We are including the *cv* calculated from all patch areas and the actual output of FRAGSTATS as well as the output of `landscapemetrics`. Interestingly, the *cv* calculated from all patches of FRAGSTATS is identical to the cv of `landscapemetrics`, but the actual result of FRAGSTATS is slightly different.

```{r}
# function to calculate coefficient of variation
cv <- function(x) {
    (sd(x) / mean(x)) * 100
}

# CV calculated from patch values of FRAGSTATS 
fragstats_calculated <- fragstats_patch_landscape %>%
    group_by(TYPE) %>%
    summarise(cv = cv(AREA)) %>%
    purrr::set_names("class", "fragstats_calculated") 

# Output of FRAGSTATS
fragstats_output <- fragstats_class_landscape %>% 
    dplyr::select(TYPE, AREA_CV) %>%
    purrr::set_names("class", "fragstats_output")

# Output of landscapemetrics
landscapemetrics_output <- lsm_c_area_cv(landscape) %>%
    dplyr::select(class, value) %>%
    purrr::set_names("class", "landscapemetrics")

fragstats <- full_join(x = fragstats_output,
                       y = fragstats_calculated, 
                       by = "class") 

cv_full <- full_join(x = fragstats, 
                     y = landscapemetrics_output,
                     by = "class")
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#1"> Show/Hide Results </button>  
<div id="1" class="collapse">

```{r}
knitr::kable(cbind(cv_full))
```

</div>

As for the *cv*, the results for the *sd* are similiar. The result calculated from all patch areas of FRAGSTATS is identical to the result of `landscapemetrics`, but not the actual result of FRAGSTATS.

```{r}
# SD calculated from patch values of FRAGSTATS
fragstats_calculated <- fragstats_patch_landscape %>%
  group_by(TYPE) %>%
  summarise(sd = sd(AREA)) %>%
  purrr::set_names("class", "fragstats_calculated") 


# Output of FRAGSTATS
fragstats_output <- fragstats_class_landscape %>%
  dplyr::select(TYPE, AREA_SD) %>%
  purrr::set_names("class", "fragstats_output")

# Output of landscapemetrics
landscapemetrics_output <- lsm_c_area_sd(landscape) %>%
  dplyr::select(class, value) %>%
  purrr::set_names("class", "landscapemetrics")

fragstats <- full_join(x = fragstats_output, 
                       y = fragstats_calculated,
                       by = "class")

sd_full <- full_join(x = fragstats,
                     y = landscapemetrics_output,
                     by = "class")
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#2"> Show/Hide Results </button>  
<div id="2" class="collapse">

```{r}
knitr::kable(cbind(cv_full))
```

</div>

## Specific differences

### GYRATE metric

According to FRAGSTATS the radius of gyration for a patch consisting of only a single cell should equal GYRATE = 0.

> [...] GYRATE = 0 when the patch consists of a single cell [...]

However, for patches containing a single cell FRAGSTATS returns a value of *GYRATE = 0.5*. In the following table, patches with an area of *area = 0.0001* consist of only one cell. 

```{r}
# Calculate patch area
fragstats_area <- fragstats_patch_landscape %>%
  dplyr::select(PID, AREA) %>%
  purrr::set_names("id", "fragstats_area")

landscapemetrics_area <- lsm_p_area(landscape) %>%
  dplyr::select(id, value) %>%
  purrr::set_names("id", "landscapemetrics_area")

# Calculate GYRATE
fragstats_gyrate <- fragstats_patch_landscape %>%
  dplyr::select(PID, GYRATE) %>%
  purrr::set_names("id", "fragstats_gyrate")

landscapemetrics_gyrate <- lsm_p_gyrate(landscape) %>%
  dplyr::select(id, value) %>%
  purrr::set_names("id", "landscapemetrics_gyrate")

fragstats <- full_join(x = fragstats_area,
                       y = fragstats_gyrate,
                       by = "id") %>%
  dplyr::filter(fragstats_area == 0.0001)

landscapemetrics <- full_join(x = landscapemetrics_area,
                              y = landscapemetrics_gyrate,
                              by = "id") %>%
  dplyr::filter(landscapemetrics_area == 0.0001)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#4"> Show/Hide Results </button>  
<div id="4" class="collapse">

```{r}
knitr::kable(cbind(fragstats, landscapemetrics))
```

</div>

Additionally, we recognized small differences for all other patches as well. However, we could not find an explanation for this difference, yet.

#### Error propagation (for metrics based on *GYRATE* metric)

##### Class level

- GYRATE_CV (`lsm_c_gyrate_cv`)
- GYRATE_MN (`lsm_c_gyrate_mn`)
- GYRATE_SD (`lsm_c_gyrate_sd`)

##### Landscape level

- GYRATE_CV (`lsm_l_gyrate_cv`)
- GYRATE_MN (`lsm_l_gyrate_mn`)
- GYRATE_SD (`lsm_l_gyrate_sd`)

### PARA metric

The documentation of FRAGSTATS defines the perimeter-area ratio the following: 

> [...] PARA equals the ratio of the patch perimeter (m) to area (m2). [...]

Contrastingly, the output of FRAGSTATS gives the result as the ratio of the patch perimeter in meters to area in hectares.

We implemented PARA as documented in the FRAGSTATS manual using square meters. Nevertheless, the differences between the softwares are only based on different units, as shown by converting the FRAGSTATS output to meters per square meters. 

```{r}
# Output of FRAGSTATS
fragstats <- fragstats_patch_landscape %>%
    dplyr::select(PID, AREA, PERIM, PARA) %>%
    purrr::set_names("id", "area", "perim", "para") %>%
    mutate(para_calculated_ha = perim / area, 
           para_calculated_m = perim / (area * 10000)) %>%
    arrange(area)

# Output of landscapemetrics
area_landscapmetrics <- lsm_p_area(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "area")

perim_landscapmetrics <- lsm_p_perim(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "perim")

para_landscapemetrics <- lsm_p_para(landscape) %>%
    dplyr::select(id, value) %>%
    purrr::set_names("id", "para")

landscapemetrics <- dplyr::full_join(x = area_landscapmetrics, 
                                     y = perim_landscapmetrics, 
                                     by = "id") %>%
    full_join(para_landscapemetrics, 
                     by = "id") %>%
    mutate(para_calculated_ha = perim / area, 
           para_calculated_m = perim / (area * 10000)) %>% 
    arrange(area)
```

<button class="btn btn-primary" data-toggle="collapse" data-target="#5"> Show/Hide Results </button>  
<div id="5" class="collapse">

```{r}
knitr::kable(cbind(fragstats, landscapemetrics))
```

</div>

#### Error propagation (for metrics based on **PARA** metric)

##### Class level
- PARA_MN (`lsm_c_para_mn`)
- PARA_SD (`lsm_c_para_sd`)

# Comparison with SDMTools

`SDMTools` (VanDerWal et al. 2014) (still available, but apparently not longer maintained) offers landscape metrics on patch and class level. However, it does not return the same results as FRAGSTATS. The main reason for this are different standard defaults (e.g. `SDMTools` always considers the global landscape boundary) and that `SDMTools` returns results in map units and not in m^2/hectar, as FRAGSTATS/`landscapemetrics`. This also explains differences between our package and `SDMTools`.

[Joseph Stachelek](https://github.com/jsta) was so nice to remind us of these issues and provided the comparison.

## Patch metrics
To get all availabel metrics on e.g. patch level with `SDMTools`, you have to make a binary landscape for every class in your landscape, perform connected components labelling on it and then calculate the patch metrics.

```{r}
library(SDMTools)

# binarize every class in the landscape and calculate patch metrics
sdmtools_result <- lapply(raster::unique(landscape), FUN = function(x){
  tmp_land <- landscape
  raster::values(tmp_land)[raster::values(tmp_land) != x] <- NA
  raster::values(tmp_land)[raster::values(tmp_land) == x] <- 1
  ccl <- SDMTools::ConnCompLabel(tmp_land)
  PatchStat(ccl)
})

# combine to one data frame
sdmtools_result <- dplyr::bind_rows(sdmtools_result, .id = "Class")
```

`landscapemetrics` offers for such tasks the function `get_patches` and for the metrics itself all of that is done internally. To get all metrics on patch level with `landscapemetrics` you could for example do:

```{r}
patch_metrics <- calculate_lsm(landscape, what = "patch")
``` 

<button class="btn btn-primary" data-toggle="collapse" data-target="#6"> Show/Hide Results </button>  
<div id="6" class="collapse">

```{r}
knitr::kable(bind_rows(sdmtools_result))
```

</div>

## References 

- McGarigal, K., SA Cushman, and E Ene. 2012. FRAGSTATS v4: Spatial Pattern Analysis Program for Categorical and Continuous Maps. Computer software program produced by the authors at the University of Massachusetts, Amherst. Available at the following website: http://www.umass.edu/landeco/research/fragstats/fragstats.html

- Jeremy VanDerWal, Lorena Falconi, Stephanie Januchowski, Luke Shoo and Collin Storlie (2014). SDMTools: Species Distribution Modelling Tools: Tools for processing data associated with species distribution modelling exercises. R package version 1.1-221. https://CRAN.R-project.org/package=SDMTools

