---
title: "Calculating landscape metrics for irregular areas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating landscape metrics for irregular areas}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

Landscape metrics are quantitative measures that describe the composition and configuration of a landscape.
In this vignette, we will use the R package **landscapemetrics** to calculate landscape metrics for irregular areas.

To get started, we first need to install and attach the required R packages. 
We will use **landscapemetrics**, **sf**, **terra**, and **tidyr** in this example:

```{r load_libraries, message = FALSE, warning = FALSE}
library(landscapemetrics)
library(sf)
library(terra)
library(tidyr)
```

Next, we need to read in our example data. 
In this case, we will use the Augusta NLCD dataset and an example set of irregular polygons provided by the **landscapemetrics** package:

```{r}
# read example data ----------------------------------------------------
augusta_nlcd = rast(landscapemetrics::augusta_nlcd)
irregular_areas <- landscapemetrics:::irregular_areas

plot(augusta_nlcd, type = "classes")
plot(st_geometry(irregular_areas), add = TRUE)
```

In the code above, we read in the Augusta NLCD dataset as a raster object and the example irregular polygons as an `sf` object.
We then plot both of these objects together to visualize our data.

We can calculate landscape metrics for **the whole landscape** using the `calculate_lsm()` function:

```{r}
# calculate lsm for the whole landscape -----------------------------------
all_landscape = calculate_lsm(augusta_nlcd, what = c("lsm_l_ent", "lsm_l_condent", 
                                                      "lsm_l_mutinf", "lsm_l_joinent", 
                                                      "lsm_l_relmutinf"))

all_landscape
```

Here, we specify the `what` argument to calculate a set of landscape metrics that include entropy, conditional entropy, mutual information, joint entropy, and relative mutual information. 
We store the calculated landscape metrics in the `all_landscape` object.

On the other hand, we can also calculate landscape metrics for irregular polygons using the `sample_lsm()` function:

```{r}
# calculate lsm for irregular polygons ------------------------------------
irr_landscapes = sample_lsm(augusta_nlcd, irregular_areas,
                            what = c("lsm_l_ent", "lsm_l_condent",
                                     "lsm_l_mutinf", "lsm_l_joinent", 
                                     "lsm_l_relmutinf"))

irr_landscapes
```

Here, we use the `sample_lsm()` function to calculate the same set of landscape metrics for our example set of irregular polygons.
We store the results in the `irr_landscapes` object.

We can connect the calculated landscape metric values to our original polygons using the `pivot_wider()` function:

```{r}
# connect values to polygons ----------------------------------------------
irr_landscapes_w = pivot_wider(irr_landscapes, 
                               id_cols = plot_id,
                               names_from = metric, 
                               values_from = value)

irregular_areas = cbind(irregular_areas, irr_landscapes_w)
```

Here, we pivot the `irr_landscapes` object from a long format to a wide format using the `pivot_wider()` function. 
We specify the `id_cols`, `names_from`, and `values_from` arguments to connect the calculated landscape metric values to our original polygons. 
We then add the calculated landscape metrics as columns to our original `irregular_areas` object using `cbind()`.

Finally, we can visualize the calculated landscape metrics using the `plot()` function:

```{r}
# viz results -------------------------------------------------------------
plot(irregular_areas["ent"])
plot(irregular_areas["condent"])
plot(irregular_areas["mutinf"])
plot(irregular_areas["joinent"])
plot(irregular_areas["relmutinf"])
```
