---
title: "New Metrics outside FRAGSTATS"
date: "`r Sys.Date()`"
output:
  github_document:
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load_libraries_hidden, eval=TRUE, echo=FALSE, message=FALSE, results='hide'}
library(landscapemetrics)
```

One of the reason to start **landscapemetrics** was also to have a collection of metrics that are not included in FRAGSTATS.
This vignette will highlight them and provide references for further reading on them.

## Landscape complexity metrics

> Nowosad J., TF Stepinski. 2018. Information-theoretical approach to measure landscape complexity. https://doi.org/10.1101/383281

The standard set of landscape metrics lacks the bona fide indicator of complexity. 
Information-theoretical framework can be applied to derive four metrics of landscape complexity: `H(x,y)` -- an overall spatio-thematic complexity metric; `H(x)` -- a thematic complexity of landscape classes; `H(y|x)` -- a configurational complexity of landscape pattern; and `I(y;x)` -- a metric disambiguating landscape pattern types having the same value of an overall complexity.
All of these metrics are implemented in **landscapemetrics**:

- [lsm_l_joinent](https://r-spatialecology.github.io/landscapemetrics/reference/lsm_l_joinent.html): an overall spatio-thematic complexity metric.
- [lsm_l_ent](https://r-spatialecology.github.io/landscapemetrics/reference/lsm_l_ent.html): a diversity (thematic complexity) of landscape classes
- [lsm_l_condent](https://r-spatialecology.github.io/landscapemetrics/reference/lsm_l_condent.html): a complexity of a landscape pattern configuration. It measures a only a geometric intricacy (configurational complexity) of a landscape pattern.
- [lsm_l_mutinf](https://r-spatialecology.github.io/landscapemetrics/reference/lsm_l_mutinf.html):  disambiguates landscape pattern types characterize by the same value of an overall complexity.
