---
title: "Utility functions"
date: "`r Sys.Date()`"
output:
  github_document:
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load_libraries_hidden, eval=TRUE, echo=FALSE, message=FALSE, results='hide'}
library(landscapemetrics)
library(ggplot2)
library(patchwork)
library(raster)
library(dplyr)
library(landscapetools)

# landscape_plot <- util_plot(landscape) + theme_void() + guides(fill = FALSE) + coord_equal()
```

## Visualisization functions

### Visualizing patches

To visualize patches in a landscape and encode each patch with an ID that can be used to compare a landscape metric with the actual landscape you can use the auxiliary visualisation function `show_patches()`:

```{r fig.height=12, fig.align="center"}
library(landscapetools)
library(patchwork)
library(landscapemetrics)
library(ggplot2)

# Create plot of the landscape
landscape_plot <- util_plot(landscape) + theme_void() + guides(fill = FALSE) + coord_equal()

# Plot landscape + landscape with labeled patches
landscape_plot  + show_patches(landscape) + plot_layout(ncol = 1)
```

### Connected labeling 

`landscapemetrics` makes internally heavy use of an connected labeling algorithm and exports an reimplementation of this algorithm (`cclabel`). You can use this function to look at patches of classes, but you can also use it to "dissect" your landscape
visually and get an impression of the spatial distribution of the classes and number of patches. The [**landscapetools**](https://github.com/ropensci/landscapetools) package offers a function to plot lists, stacks and bricks of rasters as facets. As `get_patches` returns a list with RasterLayer elements for each class, this is a convienent way
to look at patterns in your landscape:

```{r fig.width=9, fig.height=4, fig.align="center"}
# run connected labeling for landscape raster
patches <- get_patches(landscape)

# show landscape and a single facet with every patch for all classes in podlasie
util_facetplot(patches) 
```

## Show correlation

Selecting meaningful landscape metrics for your field of research is difficult,
as many landscape metrics are very abstract and the common approach is often simply 
to calculate as many as possible.

To select at the least that ones for your landscape and research question that are 
not highly correlated, you can use the function `show_correlation` to get insight
into the correlation of the metrics you calculated:

```{r fig.width=7, fig.height=7, fig.align="center"}
metrics <- calculate_metrics(landscape, what = "patch")
show_correlation(metrics, method = "pearson")
```

## Building blocks

### Get adjacencies
Adjacencies are a central part for landscape metrics, so calculating them
quick and in a flexible way is key for e.g. developing new metrics.
Hence, **landscapemetrics** exports a function that can calculate adjacencies
in any number if directions when provided with a binary matrix (NA / 1 - NA are 
cells that would be left out for looking at adjacencies).

```{r}
# calculate full adjacency matrix
get_adjacencies(landscape, 4)

# count diagonal neighbour adjacencies
diagonal_matrix <- matrix(c(1,  NA,  1,
                            NA,  0, NA,
                            1,  NA,  1), 3, 3, byrow = TRUE)
get_adjacencies(landscape, diagonal_matrix)

# equivalent with the raster package:
adj_raster <- function(landscape){
    adjacencies <- raster::adjacent(landscape, 1:raster::ncell(landscape), 4, pairs=TRUE)
    table(landscape[adjacencies[,1]], landscape[adjacencies[,2]])
}

# compare the two implementations
library(bench)
mark(
    get_adjacencies(landscape, 4),
    adj_raster(landscape),
    iterations = 100,
    relative = TRUE,
    check = FALSE
)

adj_raster(landscape) == get_adjacencies(landscape, 4)
```


### Get nearest neighbour

**landscapemetrics** implements a memory efficient and hopefully quite fast way
to calculate the nearest neighbour between classes in a raster (or matrix).

```{r}
# run connected labeling for podlasie raster
patches <- get_patches(podlasie_ccilc, 100)[[1]]

# calculate the minimum distance between patches in a landscape
min_dist <- get_nearestneighbour(patches)


# create a function that would do the same with the raster package
nearest_raster_fun <- function(patches) {
    np_class <- patches %>%
        raster::values() %>%
        unique() %>%
        na.omit() %>%
        length()
    
    points_class <- patches %>%
        raster::rasterToPoints() %>%
        tibble::as.tibble() %>%
        purrr::set_names(c("x", "y", "id"))
    
    minimum_distance <- np_class %>%
        seq_len() %>%
        purrr::map_dbl(function(patch_ij) {
            patch_focal <- dplyr::filter(points_class, id == patch_ij)
            
            patch_others <-
                dplyr::filter(points_class, id != patch_ij)
            
            minimum_distance <-
                raster::pointDistance(patch_focal[1:2],
                                      patch_others[1:2],
                                      lonlat = FALSE) %>%
                min()
        })
    
    tibble::tibble(id = unique(sort(points_class$id)),
                   distance = minimum_distance)
    
}


# compare the two implementations
library(bench)
mark(
    get_nearestneighbour(patches),
    nearest_raster_fun(patches),
    iterations = 100,
    relative = TRUE,
    check = FALSE
)

all(round(get_nearestneighbour(patches)$distance, 9) ==
        round(nearest_raster_fun(patches)$distance, 9)) 
```
