---
title: "Communicate landscapemetrics results"
date: "`r Sys.Date()`"
output:
  github_document:
    html_preview: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load_libraries_hidden, eval=TRUE, echo=FALSE, message=FALSE, results='hide'}
library(landscapemetrics)
library(ggplot2)
library(patchwork)
library(raster)
library(dplyr)
library(landscapetools)

# landscape_plot <- util_plot(landscape) + theme_void() + guides(fill = FALSE) + coord_equal()
```

## Visualizing patches

To visualize patches in a landscape and encode each patch with an ID that can be used to compare a landscape metric with the actual landscape you can use the auxiliary visualisation function `show_patches()`:

```{r fig.height=12, fig.align="center"}
library(landscapetools)
library(patchwork)
library(landscapemetrics)
library(ggplot2)

# Create plot of the landscape
landscape_plot <- util_plot(landscape) + theme_void() + guides(fill = FALSE) + coord_equal()

# Plot landscape + landscape with labeled patches
landscape_plot  + show_patches(landscape) + plot_layout(ncol = 1)
```

## Connected labeling 

`landscapemetrics` makes internally heavy use of an implementatian of the connected labeling algorithm by Thell Fowler (https://github.com/Thell/ccloutline) and exports an reimplementation of this algorithm (`cclabel`). You can use this function to look at patches of classes, but you can also use it to "dissect" your landscape
visually and get an impression of the spatial distribution of the classes and number of patches. The [**landscapetools**](https://github.com/ropensci/landscapetools) package offers a function to plot lists, stacks and bricks of rasters as facets. As `cclabel` returns a list with RasterLayer elements for each class, this is a convienent way
to look at patterns in your landscape:

```{r fig.width=9, fig.height=4, fig.align="center"}
# run connected labeling for podlasie raster
patches <- cclabel(landscape)

# show landscape and a single facet with every patch for all classes in podlasie
util_facetplot(patches) 
```

## Show correlation

Selecting meaningful landscape metrics for your field of research is difficult,
as many landscape metrics are very abstract and the common approach is often simply 
to calculate as many as possible.

To select at the least that ones for your landscape and research question that are 
not highly correlated, you can use the function `show_correlation` to get insight
into the correlation of the metrics you calculated:

```{r fig.width=7, fig.height=7, fig.align="center"}
metrics <- lsm_calculate(landscape, what = "patch")
show_correlation(metrics, level = "patch", method = "pearson")
```
